---
title: "Assignment #1"
format: html
---

## Assignment 1

You only need to write lines of code for each question. When answering questions that ask you to identify or interpret something, the length of your response doesn’t matter. For example, if the answer is just ‘yes,’ ‘no,’ or a number, you can just give that answer without adding anything else.

We will go through comparable code and concepts in the live learning session. If you run into trouble, start by using the help help() function in R, to get information about the datasets and function in question. The internet is also a great resource when coding (though note that no outside searches are required by the assignment!). If you do incorporate code from the internet, please cite the source within your code (providing a URL is sufficient).

Please bring questions that you cannot work out on your own to office hours, work periods or share with your peers on Slack. We will work with you through the issue.

You will need to install PLINK and run the analyses. Please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md). PLINK is a free, open-source whole genome association analysis toolset, designed to perform a range of basic, large-scale analyses in a computationally efficient manner.

```{r setup, include=FALSE}
# Set up a consistent path for all chunks of code
knitr::opts_knit$set(root.dir = normalizePath("../../"))

library(data.table)
library(ggplot2)
library(seqminer)
library(HardyWeinberg)
library(dplyr)
```


#### Question 1: Data inspection

Before fitting any models, it is essential to understand the data. Use R or bash code to answer the following questions about the `gwa.qc.A1.fam`, `gwa.qc.A1.bim`, and `gwa.qc.A1.bed` files, available at the following Google Drive link: <https://drive.google.com/drive/folders/11meVqGCY5yAyI1fh-fAlMEXQt0VmRGuz?usp=drive_link>. Please download all three files from this link and place them in `02_activities/data/`.

(i) Read the .fam file. How many samples does the dataset contain?

```{bash}         
#Read the .fam file
head ./02_activities/data/gwa.qc.A1.fam

#The dataset contains 10 samples.
```

(ii) What is the 'variable type' of the response variable (i.e.Continuous or binary)?

```{bash}         
head ./02_activities/data/gwa.qc.A1.fam

#The response variable is continuous.
```

(iii) Read the .bim file. How many SNPs does the dataset contain?

```{bash}         
head ./02_activities/data/gwa.qc.A1.bim

#There are 10 SNPs in the dataset.
```

#### Question 2: Allele Frequency Estimation

(i) Load the genotype matrix for SNPs rs1861, rs3813199, rs3128342, and rs11804831 using additive coding. What are the allele frequencies (AFs) for these four SNPs?

```{bash}         
#Create SNP list file with 4 SNPs: rs1861, rs3813199, rs3128342, and rs11804831
printf "%s\n" rs1861 rs3813199 rs3128342 rs11804831 > ./02_activities/data/A1snplist.txt

#Subset SNPs and output allele frequencies
plink2 --bfile ./02_activities/data/gwa.qc.A1 \
      --extract ./02_activities/data/A1snplist.txt \
      --recode A\
      --freq \
      --out ./02_activities/data/gwa_qc_A1_freq
```
```{r}
#View allele frequencies
A1freq <- fread("./02_activities/data/gwa_qc_A1_freq.afreq")
head(A1freq)

#The allele frequencies for rs1861, rs3813199, rs3128342, and rs11804831 are as follows: 0.0539859, 0.0569126, 0.3051210, and 0.1543410.
```

(ii) What are the minor allele frequencies (MAFs) for these four SNPs?

```{r}         
A1freq$MAF <- pmin(A1freq$ALT_FREQS, 1 - A1freq$ALT_FREQS)

A1freq[, .(ID, AF = ALT_FREQS, MAF)]

#The minor allele frequencies for rs1861, rs3813199, rs3128342, and rs11804831 are as follows: 0.0539859, 0.0569126, 0.3051210, and 0.1543410.
```


#### Question 3: Hardy–Weinberg Equilibrium (HWE) Test

(i) Conduct the Hardy–Weinberg Equilibrium (HWE) test for all SNPs in the .bim file. Then, load the file containing the HWE p-value results and display the first few rows of the resulting data frame.

```{bash}         
plink2 --bfile ./02_activities/data/gwa.qc.A1 --hardy --out ./02_activities/data/gwa_qc_A1_hwe
```

```{r}
A1hwe <- fread("./02_activities/data/gwa_qc_A1_hwe.hardy")
head(A1hwe)
```

(ii) What are the HWE p-values for SNPs rs1861, rs3813199, rs3128342, and rs11804831?

```{bash}         
#Subset SNPs and output allele frequencies
plink2 --bfile ./02_activities/data/gwa.qc.A1 \
      --extract ./02_activities/data/A1snplist.txt \
      --hardy \
      --out ./02_activities/data/gwa_qc_A1_hwe_snplist
```

```{r}
#View allele frequencies
A1hwesnplist <- fread("./02_activities/data/gwa_qc_A1_hwe_snplist.hardy")
head(A1hwesnplist)

#The HWE p-values for rs1861, rs3813199, rs3128342, and rs11804831 are the following: 0.274719, 1.000000, 0.330273, and 0.113354.
```

#### Question 4: Genetic Association Test

(i) Conduct a linear regression to test the association between SNP rs1861 and the phenotype. What is the p-value?

```{bash}
plink2 --bfile ./02_activities/data/gwa.qc.A1 \
      --extract ./02_activities/data/A1snplist.txt \
      --recode A \
      --out ./02_activities/data/geno.A1.additive
```

```{r}
# Load genotype data
geno_A1_subset <- fread("./02_activities/data/geno.A1.additive.raw")
geno_A1_subset=na.omit(geno_A1_subset)
geno_A1_subset$PHENOTYPE=geno_A1_subset$PHENOTYPE-1
head(geno_A1_subset)

#Subset to just rs1861
geno_rs1861 <- geno_A1_subset %>%
  select(FID, IID, PHENOTYPE, rs1861_C) %>%
  na.omit()

#Conduct a linear regression model between rs1861 and the phenotype
lm_rs1861 <- lm(PHENOTYPE ~ rs1861_C, data = geno_rs1861)
summary(lm_rs1861)

#The p-value is less than 2 × 10⁻¹⁶. There is a statistically significant relationship between rs1861 and the phenotype.
```

(ii) How would you interpret the beta coefficient from this regression?

```         
# We performed a linear regression to test the association between genotype rs1861 and the continuous phenotype. The beta-coefficient is 0.96698. This means that for every additional C allele an individual has for rs1861, their phenotype increases by 0.96698 units.
```

(iii) Plot the scatterplot of phenotype versus the genotype of SNP rs1861. Add the regression line to the plot.

```{r}         
# Create a new data frame containing a sequence of genotype values
 geno_range <- data.frame(
   rs1861_C = seq(0, 2, length.out = 100)
 )
# Use the fitted linear regression model to predict phenotype probability
# for each genotype value in geno_range
 geno_range$predicted_prob <- predict(lm_rs1861, newdata = geno_range, type = "response")

# Start from the observed genotype dataset
count_data <- geno_A1_subset %>%
# Remove rows with missing genotype or missing phenotype
  filter(!is.na(rs1861_C), !is.na(PHENOTYPE)) %>%
 # Group data by genotype value and phenotype category 
  group_by(rs1861_C, PHENOTYPE) %>%
# Count how many observations fall into each genotype-phenotype combination
  summarise(count = n(), .groups = "drop")

# Start building the plot
p1=ggplot() +
# Add points showing observed genotype-phenotype combinations
  geom_point(
    # Use the summarized count_data for the points
    data = count_data,
    # Map genotype to x-axis, phenotype to y-axis, and count to point size
    aes(x = rs1861_C, y = PHENOTYPE, size = count),
    color = "blue"
  ) +
  # Add a regression line showing predicted probability across genotype values
  geom_line(# Use the prediction data frame for the line
    data = geno_range,
    aes(x = rs1861_C, y = predicted_prob),
    color = "red",
    # Set line thickness
    size = 1.2
  ) +
  # Control the size range of the points
  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Linear Regression: PHENOTYPE ~ Genotype",
    x = "Genotype (0/1/2)",
    y = "Phenotype ",
    size = "Count"
  ) +
  # Restrict the visible x- and y-axis range
  coord_cartesian(xlim = c(-0.5, 2.5), ylim = c(-0.1, 1.1)) +
   # Apply a minimal theme for a clean appearance
  theme_minimal()

print(p1)
```

(iv) Convert the genotype coding for rs1861 to recessive coding.

``` {r}  
geno_A1_subset_rec <- geno_A1_subset # make a copy

#change into recessive coding
geno_A1_subset_rec[, 7:ncol(geno_A1_subset_rec)] <- as.data.frame(
  lapply(geno_A1_subset[, 7:ncol(geno_A1_subset)], function(x) ifelse(x == 2, 1, 0))
)

#Subset to just rs1861
geno_rs1861_rec <- geno_A1_subset_rec %>%
  select(FID, IID, PHENOTYPE, rs1861_C) %>%
  na.omit()

```

(v) Conduct a linear regression to test the association between the recessive-coded rs1861 and the phenotype. What is the p-value?

``` {r}        
#Conduct a linear regression model between rs1861 and the phenotype under recessive coding
lm_rs1861_rec <- lm(PHENOTYPE ~ rs1861_C, data = geno_rs1861_rec)
summary(lm_rs1861_rec)

#The p-value is less than 2 × 10⁻¹⁶. There is a statistically significant relationship between rs1861 and the phenotype.
```

(vi) Plot the scatterplot of phenotype versus the recessive-coded genotype of rs1861. Add the regression line to the plot.

```{r}         
# Create a new data frame containing a sequence of genotype values
 geno_range_rec <- data.frame(
   rs1861_C = seq(0, 2, length.out = 100)
 )
# Use the fitted linear regression model to predict phenotype probability
# for each genotype value in geno_range
 geno_range_rec$predicted_prob <- predict(lm_rs1861_rec, newdata = geno_range_rec, type = "response")

# Start from the observed genotype dataset
count_data_rec <- geno_A1_subset_rec %>%
# Remove rows with missing genotype or missing phenotype
  filter(!is.na(rs1861_C), !is.na(PHENOTYPE)) %>%
 # Group data by genotype value and phenotype category 
  group_by(rs1861_C, PHENOTYPE) %>%
# Count how many observations fall into each genotype-phenotype combination
  summarise(count = n(), .groups = "drop")

# Start building the plot
p2=ggplot() +
# Add points showing observed genotype-phenotype combinations
  geom_point(
    # Use the summarized count_data for the points
    data = count_data_rec,
    # Map genotype to x-axis, phenotype to y-axis, and count to point size
    aes(x = rs1861_C, y = PHENOTYPE, size = count),
    color = "blue"
  ) +
  # Add a regression line showing predicted probability across genotype values
  geom_line(# Use the prediction data frame for the line
    data = geno_range_rec,
    aes(x = rs1861_C, y = predicted_prob),
    color = "red",
    # Set line thickness
    size = 1.2
  ) +
  # Control the size range of the points
  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Linear Regression: PHENOTYPE ~ Genotype",
    x = "Genotype (0/1)",
    y = "Phenotype ",
    size = "Count"
  ) +
  # Restrict the visible x- and y-axis range
  coord_cartesian(xlim = c(-0.5,1.5), ylim = c(-0.1, 1.1)) +
  scale_x_continuous(breaks = c(0, 1))+
   # Apply a minimal theme for a clean appearance
  theme_minimal()

print(p2)
```

(vii) Which model fits better? Justify your answer.

```{r}
AIC(lm_rs1861, lm_rs1861_rec)
```

```         
# The AIC is higher in the recessive model (AIC=11000.02), indicating that the additive model (AIC=10988.06) may fit better slightly better overall by capturing a more gradual allele-phenotype effect.

# However, it is also important to note that there are not a lot of individuals with the homozygous variant genotype in this dataset. As a result, that may be why the recessive model and additive models fit similarly.
```

### Criteria

| Criteria | Complete | Incomplete |
|------------------------|------------------------|------------------------|
| **Data Inspection** | Correct sample/SNP counts and variable type identified. | Missing or incorrect counts or variable type. |
| **Allele Frequency Estimation** | Correct allele and minor allele frequencies computed. | Frequencies missing or wrong. |
| **Hardy–Weinberg Equilibrium Test** | Correct PLINK command and p-value extraction in R. | PLINK command or extraction incorrect/missing. |
| **Genetic Association Test** | Correct regressions, plots, coding, and interpretation. | Regression, plots, or interpretation missing/incomplete. |

### Submission Information

📌 Please review our [Assignment Submission Guide](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md) for detailed instructions on how to format, branch, and submit your work. Following these guidelines is crucial for your submissions to be evaluated correctly.

#### Note:

If you like, you may collaborate with others in the cohort. If you choose to do so, please indicate with whom you have worked with in your pull request by tagging their GitHub username. Separate submissions are required.

------------------------------------------------------------------------

#### Submission Parameters

-   Submission Due Date: `11:59 PM – 16/03/2026`

-   Branch name for your repo should be: `assignment-1`

-   What to submit for this assignment:

    -   Populate this Quarto document (`assignment_1.qmd`).
    -   Render the document with Quarto: `quarto render assignment_1.qmd`.
    -   Submit both `assignment_1.qmd` and the rendered HTML file `assignment_1.html` in your pull request.

-   What the pull request link should look like for this assignment: `https://github.com/<your_github_username>/gen_data/pull/<pr_id>`

    -   Open a private window in your browser. Copy and paste the link to your pull request into the address bar. Make sure you can see your pull request properly. This helps the technical facilitator and learning support team review your submission easily.

------------------------------------------------------------------------

Checklist:

-   Created a branch with the correct naming convention.
-   Ensured that the repository is public.
-   Reviewed the PR description guidelines and adhered to them.
-   Verified that the link is accessible in a private browser window.
-   Confirmed that both `assignment_1.qmd` and `assignment_1.html` are included in the pull request.

If you encounter any difficulties or have questions, please don't hesitate to reach out to our team via our Slack help channel. Our technical facilitators and learning support team are here to help you navigate any challenges.
